---
title: "Modeling Runners' Times in the Cherry Blossom Race"
author: "Justin Howard, Stuart Miller, Paul Adams"
date: "September 22, 2020"
lang: en-US
class: man
# figsintext: true
numbersections: true
encoding: UTF-8
output:
  bookdown::pdf_document2:
     citation_package: natbib
     keep_tex: true
     toc: false
header-includes:
   - \usepackage{amsmath}
   - \usepackage{gensymb}
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{setspace}
   - \usepackage{hyperref}
   - \onehalfspacing
editor_options: 
  chunk_output_type: console
---

# Introduction

The internet is a vast open resource for data, but internet data can be messy and difficult to wrangle. 
In this case study we collect data on participants of the Cherry Blossom Race from the event website\footnotemark.
We find that there are a number of issues present in the data, 
 including inconsistent web addressing formatting, inconsistent formatting, and missing data. 
We collect and clean this data to provide the Men's and Women's information in a format suitable for analysis.


\footnotetext{See http://www.cherryblossom.org/.}


# Methods

## Data 


### Data Collection

In this case study, we collected data on participants in the Cherry Blossom Race. We scraped the data for Men and Woman from 1999 to 2012. The data was stored under the base address in the following form, where the year is between 1999 and 2012 and page is a specific page name.

```
http://www.cherryblossom.org/results/<year>/<page>
```

The specific addresses where the data is stored are given in table 1. While the structure follows `results/<year>` consistently, the page naming for each year is not consistent. In some cases, the names for the men's and women's pages are `men` and `women`, repectively. In other cases, a code is used such as `cb99m` and `cb99f` in 1999 for `men` and `women` repectively.
\newpage
**Table 1. Web Pages**

| Year | Men's Pages                      | Women's Pages             |
|------|----------------------------------|---------------------------|
| 1999 | `results/1999/cb99m.html`        | `results/1999/cb99f.html` |
| 2000 | `results/2000/Cb003m.htm`        | `results/2000/Cb003f.htm` |
| 2001 | `results/2001/oof_m.html`        | `results/2001/oof_f.html` |
| 2002 | `results/2002/oofm.htm`          | `results/2002/ooff.htm`   |
| 2003 | `results/2003/CB03-M.HTM`        | `results/2003/CB03-F.HTM` |
| 2004 | `results/2004/men.htm`           | `results/2004/women.htm`  |
| 2005 | `results/2005/CB05-M.htm`        | `results/2005/CB05-F.htm` |
| 2006 | `results/2006/men.htm`           | `results/2006/women.htm`  |
| 2007 | `results/2007/men.htm`           | `results/2007/women.ht`   |
| 2008 | `results/2008/men.htm`           | `results/2008/women.htm`  |
| 2009 | `results/2009/09cucb-M.htm`      | `results/2009/09cucb-F.htm`      |
| 2010 | `results/2010/2010cucb10m-m.htm` | `results/2010/2010cucb10m-F.htm` |
| 2011 | `results/2011/2011cucb10m-m.htm` | `results/2011/2011cucb10m-F.htm` |
| 2012 | `results/2012/2012cucb10m-m.htm` | `results/2012/2012cucb10m-F.htm` |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = F)
```

### Data Cleaning

\small
*This section refers to code in Appendix A*

\normalsize

```{r}

library(pacman)
p_load(XML, rnaturalearth,rnaturalearthdata,countrycode,tidyverse,ggplot2,ggthemes,viridis,rgeos,usmap, sqldf, maps)
ubase = "http://www.cherryblossom.org/"
menURLs = 
  c("results/1999/cb99m.html", "results/2000/Cb003m.htm", "results/2001/oof_m.html",
    "results/2002/oofm.htm", "results/2003/CB03-M.HTM",
    "results/2004/men.htm", "results/2005/CB05-M.htm", 
    "results/2006/men.htm", "results/2007/men.htm", 
    "results/2008/men.htm", "results/2009/09cucb-M.htm",
    "results/2010/2010cucb10m-m.htm", 
    "results/2011/2011cucb10m-m.htm",
    "results/2012/2012cucb10m-m.htm")

urls = paste(ubase, menURLs, sep = "")

extractResTable =
  # takes a list of websites from the cherry blossom race
  # a list of years corresponding to the year the result is for
  # and the gender of the participant
  # Retrieve data from web site, 
  # find the preformatted text,
  # and write lines or return as a character vector.
  # returns a list of strings corrsponding to lines in the web url
  function(url = "http://www.cherryblossom.org/results/2009/09cucb-F.htm",
           year = 1999, sex = "male", file = NULL)
  {
    doc = htmlParse(url)

    if (year == 2000) {
      # Get preformatted text from 4th font element
      # The top file is ill formed so the <pre> search doesn't work.
      ff = getNodeSet(doc, "//font")
      txt = xmlValue(ff[[4]])
      els = strsplit(txt, "\r\n")[[1]]
    }
    else if (year == 2009 & sex == "male") {
      # Get preformatted text from <div class="Section1"> element
      # Each line of results is in a <pre> element
      div1 = getNodeSet(doc, "//div[@class='Section1']")
      pres = getNodeSet(div1[[1]], "//pre")
      els = sapply(pres, xmlValue)
      els = gsub("Â", "", els)
    }
    else if (year == 1999) {
      # Get preformatted text from <pre> elements
      pres = getNodeSet(doc, "//pre")
      txt = xmlValue(pres[[1]])
      els = strsplit(txt, "\n")[[1]]   
    } 
    else {
      # Get preformatted text from <pre> elements
      pres = getNodeSet(doc, "//pre")
      txt = xmlValue(pres[[1]])
      els = strsplit(txt, "\r\n")[[1]]   
    } 
    if (is.null(file)) return(els)
    # Write the lines as a text file.
    writeLines(els, con = file)
  }


years = 1999:2012
menTables = mapply(extractResTable, url = urls, year = years)
names(menTables) = years
sapply(menTables, length)

save(menTables, file = "CBMenTextTables.rda")
```

The data are successfully scraped from the urls that are given and we observe that the number of participants in the Cherry Blossom race has increased steadily from 1999 to 2012. We can use the `extractResTable` function to exptract all of the results, or to specify a single set of results.


```{r, echo = TRUE, inlcude = T, results='asis'}
men_2009<-extractResTable(url = urls[11], year = 2009, sex = 'male')
men_2009[1:10]
```
\normalsize

We observe the formatting of the text data, such as the use of the `=` character can be used to transform the lines of text into a a matrix. We will used this pattern to identify the boundary between the headering and the body of the data. Once we defined the column names, we were free to use the text structure, such as the spacing of column names and data, to identify datapoints. We started with age, which denoted by the `"ag"` identifier. After we sucessfully used the structure of the text data to identify an index location for `"ag"` data points. We expanded the search to inlcude  the spacing of text data to identify other data points. The spacing indices were used in a funtion that compiled a matrix of data points across the entire list of tables. 

A function `findColLocs` was defined to use these space indices to locate the begin points for data throughout the extracted text data. `selectCols` is a tool that uses `findColLocs` to locate the data points for each column throughout the text for all of the race years. Our functions extracted almost all of the age data fomr the text, which aided in the identification errors within the dataset. To generalize this method, we defined a function, `extractVariables`, that combined the various processes we used to preprocess the data into a more machine readable format. 

We now have a list of matrices. The original text data are in a much more machine readable format. Our number of rows for each matrix is equal to the number of original rows, minus the original text header. To perform statistical analyses on the numeric data, we must perform a datatype transformation.

There were anomalies in the data for the year 2003. After an examination of the original data table, we found that the spacing in the year 2003 data is not the same as the other years. We adjusted our `selectCols` function to better capture the data for 2003. We also observe anomalies in the formatting of the year 2001. We introduced fixes and re-scraped the data once more to form the list of matrices called `menResMat`.

```{r}
equals_idx<- grep('^===', men_2009)
equals_idx

equals_idx2<- substr(men_2009, 1,3)
which(equals_idx2 == '===')

spacerRow<- men_2009[equals_idx]
headerRow<- men_2009[equals_idx -1]
body<- men_2009[-(1:equals_idx)] # working with 2009 body
```
```{r}
head(body)
```

```{r}
headerRow<- tolower(headerRow)
headerRow
```

```{r, inlcude = FALSE, results= 'hide'}
# looks for first match of given sequence
ageStart<- regexec('ag', headerRow, fixed = TRUE)
#ageStart
```

```{r}
age<- substr(body, start = 49, stop = 50)
head(age)
```
```{r}
summary(as.numeric(age))
```



```{r}
#gregexpr looks for all matches of a given sequence
blankLocs<- gregexpr(' ', spacerRow)
blankLocs
```
```{r}
searchLocs<- c(0,blankLocs[[1]])
searchLocs
```



```{r}
values<- mapply(substr, list(body), start = searchLocs[-length(searchLocs)] +1, stop = searchLocs[-1] -1)
length(values)
head(values)
```



```{r}
findColLocs<- function(spacerRow){
  spaceLocs<- gregexpr(' ', spacerRow)[[1]]
  rowLength<- nchar(spacerRow)
  
  if (substring(spacerRow, rowLength, rowLength) != ' ')
    return( c(0, spaceLocs, rowLength +1))
  else return(c(0,spaceLocs))
}

locs_2012<- findColLocs(spacerRow)
length(locs_2012)
locs_2012
```



```{r}
selectCols = 
function(colNames, headerRow, searchLocs) 
{
  sapply(colNames, 
         function(name, headerRow, searchLocs)
         {
           startPos = regexpr(name, headerRow)[[1]]
           if (startPos == -1) 
             return( c(NA, NA) )
    
           index = sum(startPos >= searchLocs)
           c(searchLocs[index] + 1, searchLocs[index + 1] - 1)
         },
         headerRow = headerRow, searchLocs = searchLocs )
}
```



```{r}
searchLocs = findColLocs(spacerRow)
ageLoc = selectCols("ag", headerRow, searchLocs) 
ages = mapply(substr, list(body), 
              start = ageLoc[1,], stop = ageLoc[2, ])
summary(as.numeric(ages))
```



```{r}
shortColNames = c("name", "home", "ag", "gun", "net", "time")
locCols = selectCols(shortColNames, headerRow, searchLocs)

#mapply forms a matrix after completion
Values = mapply(substr, list(body), start = locCols[1, ], 
                stop = locCols[2, ])

class(Values)

colnames(Values) = shortColNames
head(Values)
tail(Values)[ , 1:3]
```



```{r}
extractVariables = 
  function(file, varNames =c("name", "home", "ag", "gun",
                             "net", "time"))
{
       # Find the index of the row with =s
  eqIndex = grep("^===", file)
       # Extract the two key rows and the data
  spacerRow = file[eqIndex] 
  headerRow = tolower(file[ eqIndex - 1 ])
  body = file[ -(1 : eqIndex) ]
       
       # Obtain the starting and ending positions of variables
  searchLocs = findColLocs(spacerRow)
  locCols = selectCols(varNames, headerRow, searchLocs)

  Values = mapply(substr, list(body), start = locCols[1, ], 
                  stop = locCols[2, ])
  colnames(Values) = varNames
  
  invisible(Values)
}
```
```{r}

years = 1999:2012
menTables = mapply(extractResTable, url = urls, year = years)
names(menTables) = years
```



```{r}
menResMat = lapply(menTables, extractVariables)
length(menResMat)

sapply(menResMat, nrow)
```



```{r}
age = as.numeric(menResMat[['2012']][ , 'ag'])

tail(age)

age = sapply(menResMat,
             function(x) as.numeric(x[ , 'ag']))

```

```{r, include = TRUE}
boxplot(age, ylab = "Age", xlab = "Year")
```
```{r}
selectCols = function(shortColNames, headerRow, searchLocs) {
  sapply(shortColNames, function(shortName, headerRow, searchLocs){
    startPos = regexpr(shortName, headerRow)[[1]]
    if (startPos == -1) return( c(NA, NA) )
    index = sum(startPos >= searchLocs)
    c(searchLocs[index] + 1, searchLocs[index + 1])
  }, headerRow = headerRow, searchLocs = searchLocs )
}
```



```{r}
age2001 = age[["2001"]]

grep("^===", menResMat['2001'])

badAgeIndex = which(is.na(age2001)) + 5

extractVariables = 
function(file, varNames =c("name", "home", "ag", "gun",
                           "net", "time"))
{
  
  # Find the index of the row with =s
  eqIndex = grep("^===", file)
  # Extract the two key rows and the data 
  spacerRow = file[eqIndex] 
  headerRow = tolower(file[ eqIndex - 1 ])
  body = file[ -(1 : eqIndex) ]
       # Remove footnotes and blank rows
  footnotes = grep("^[[:blank:]]*(\\*|\\#)", body)
  if ( length(footnotes) > 0 ) body = body[ -footnotes ]
  blanks = grep("^[[:blank:]]*$", body)
  if (length(blanks) > 0 ) body = body[ -blanks ]
  
  
  # Obtain the starting and ending positions of variables   
  searchLocs = findColLocs(spacerRow)
  locCols = selectCols(varNames, headerRow, searchLocs)
  
  Values = mapply(substr, list(body), start = locCols[1, ], 
                  stop = locCols[2, ])
  colnames(Values) = varNames
  
  return(Values)
}

menResMat = lapply(menTables, extractVariables)
```

After these anomalies were corrected, we reformatted the datatypes. The `"ag"` data was changed to a numeric format for analysis, which revealed *more* anomalies. We observed suspiciously young ages between the years 2001 and 2003. These ages were revealed to be *erroneous* upon closer examination. These values were removed from the dataset.

To make the remaining analyses easier, we transformed the data structure from a list of matrices into a dataframe. Combining the data into a single dataframe facilitated the cleaning of the columns containing the time data for each runner. These values were stored as characters and were in an HR:MIN:SEC format. We can separate the values and convert them all into a standard time format, minutes. We defined a function that converts the time into a minute format. 

The end result was a dataframe called menDF. With this process complete, we simply re-used the functions on the women's data. 

```{r}
age<- sapply(menResMat, function(x) as.numeric(x[, 'ag']))
```

```{r}
sapply(age, function(x) sum(is.na(x)))
```

```{r}
boxplot(age, ylab = "Age", xlab = "Year")
```

```{r}
young<- menResMat[['2001']][,'ag'] < 10
menResMat[['2001']]<- menResMat[['2001']][-c(young),]
y2002<- menResMat[['2002']][,'ag'] < 10
y2003<- menResMat[['2003']][,'ag'] < 10
menResMat[['2002']]<- menResMat[['2002']][-c(y2002),]
menResMat[['2003']]<- menResMat[['2003']][-c(y2003),]
y2002<- menResMat[['2002']][,'ag'] < 10
y2003<- menResMat[['2003']][,'ag'] < 10
```

```{r}
menResDF = lapply(menResMat, data.frame)
```
```{r}
col_1999 = rep(1999, nrow(menResDF$`1999`))

col_2000 = rep(2000, nrow(menResDF$`2000`))

col_2001 = rep(2001, nrow(menResDF$`2001`))

col_2002 = rep(2002, nrow(menResDF$`2002`))

col_2003 = rep(2003, nrow(menResDF$`2003`))

col_2004 = rep(2004, nrow(menResDF$`2004`))

col_2005 = rep(2005, nrow(menResDF$`2005`))

col_2006 = rep(2006, nrow(menResDF$`2006`))

col_2007 = rep(2007, nrow(menResDF$`2007`))

col_2008 = rep(2008, nrow(menResDF$`2008`))

col_2009 = rep(2009, nrow(menResDF$`2009`))

col_2010 = rep(2010, nrow(menResDF$`2010`))

col_2011 = rep(2011, nrow(menResDF$`2011`))

col_2012 = rep(2012, nrow(menResDF$`2012`))

cols = c(col_1999, col_2000, col_2001, col_2002, col_2003, col_2004, col_2005, col_2006, col_2007, col_2008, col_2009, col_2010, col_2011, col_2012)
```

```{r}
menResFinal = data.frame()
for (i in 1:length(names(menResDF))) {
  
  df<- menResDF[[i]]
  menResFinal = rbind(menResFinal, df)
}
menResFinal$Date<- cols
```


```{r}
menResFinal$ag = as.numeric(menResFinal$ag)
sum(is.na(menResFinal$ag))
```


```{r, include = FALSE}
times = menResFinal$time
timePieces = strsplit(as.character(times), ":")

tail(timePieces)

timePieces = sapply(timePieces, as.numeric)
head(timePieces)
```

```{r}
runTime = sapply(timePieces, 
                 function(x) {
                   if (length(x) == 2) x[1] + x[2]/60
                   else 60*x[1] + x[2] + x[3]/60
                 })

summary(runTime)
```
```{r}
convertTime = function(time) {
  timePieces = strsplit(time, ":")
  timePieces = sapply(timePieces, as.numeric)
  sapply(timePieces, function(x) {
                      if (length(x) == 2) x[1] + x[2]/60
                      else 60*x[1] + x[2] + x[3]/60
                      })
}
```

```{r, include = FALSE}
createDF = function(Res, year, sex) 
{
  # Determine which time to use
  if ( !is.na(Res[1, 'net']) ) useTime = Res[ , 'net']
  else if ( !is.na(Res[1, 'gun']) ) useTime = Res[ , 'gun']
  else useTime = Res[ , 'time']
  
  # Remove # and * and blanks from time
  useTime = gsub("[#\\*[:blank:]]", "", useTime)
  runTime = convertTime(useTime[ useTime != "" ])
  
  # Drop rows with no time
  Res = Res[ useTime != "", ]
  
  Results = data.frame(year = rep(year, nrow(Res)),
                       sex = rep(sex, nrow(Res)),
                       name = Res[ , 'name'], home = Res[ , 'home'],
                       age = as.numeric(Res[, 'ag']), 
                       runTime = runTime,
                       stringsAsFactors = FALSE)
  invisible(Results)
}

menDF = mapply(createDF, menResMat, year = 1999:2012,
               sex = rep("M", 14), SIMPLIFY = FALSE)
```


```{r}
sapply(menDF, function(x) sum(is.na(x$runTime)))

separatorIdx = grep("^===", menTables[["2006"]])
separatorRow = menTables[['2006']][separatorIdx]
separatorRowX = paste(substring(separatorRow, 1, 63), " ", 
                      substring(separatorRow, 65, nchar(separatorRow)), 
                      sep = "")
menTables[['2006']][separatorIdx] = separatorRowX

menResMat = sapply(menTables, extractVariables)
menDF = mapply(createDF, menResMat, year = 1999:2012,
               sex = rep("M", 14), SIMPLIFY = FALSE)
```



```{r}
womenURLs = 
  c("results/1999/cb99f.html", "results/2000/Cb003f.htm", "results/2001/oof_f.html",
    "results/2002/ooff.htm", "results/2003/CB03-F.HTM",
    "results/2004/women.htm", "results/2005/CB05-F.htm", 
    "results/2006/women.htm", "results/2007/women.htm", 
    "results/2008/women.htm", "results/2009/09cucb-F.htm",
    "results/2010/2010cucb10m-F.htm", 
    "results/2011/2011cucb10m-F.htm",
    "results/2012/2012cucb10m-F.htm")

years = 1999:2012
urls = paste(ubase, womenURLs, sep = "")
womenTables = mapply(extractResTable, url = urls, year = years, sex='female')
names(womenTables) = years
sapply(womenTables, length)
```


```{r, inlcude = FALSE}
separatorIdx = grep("^===", womenTables[["2006"]])
separatorRow = womenTables[['2006']][separatorIdx]
separatorRowX = paste(substring(separatorRow, 1, 63), " ",
                      substring(separatorRow, 65, nchar(separatorRow)),
                      sep = "")
womenTables[['2006']][separatorIdx] = separatorRowX

womenTables[['2001']][2:3] = womenTables[['2002']][2:3]

womenResMat = sapply(womenTables, extractVariables)
womenDF = mapply(createDF, womenResMat, year = 1999:2012,
               sex = rep("W", 14), SIMPLIFY = FALSE)
```



```{r}
sapply(womenDF, function(x) sum(is.na(x$age)))
```



```{r}
sapply(womenDF, function(x) sum(is.na(x$runTime)))
```



```{r}
menDF_complete = na.omit(menDF)
womenDF_complete = na.omit(womenDF)
```



```{r}
cherryblossom = data.frame()
for (i in 1:14) {
  
  df<- menDF_complete[[i]]
  cherryblossom = rbind(cherryblossom, df)
}
head(cherryblossom)
```

```{r}
for (i in 1:14) {
  
  df<- womenDF_complete[[i]]
  cherryblossom = rbind(cherryblossom, df)
}
tail(cherryblossom)
```
```{r}
sum(is.na(cherryblossom))
```

In preparation for further analysis, the values for 'home' needed to be updated. First, 12 runners were identified as having at least one iteration of a bad address. Based on state codes provided with those addresses, alternative addresses used by those runners in other races were applied in place of the bad addresses.

Additionally, because most of the addresses were not in standard format, we converted home names to ISO standards for national naming. For example, we converted 'Rep Of S.africa' to the iso3166-recognized 'South Africa'. The names were originally given ISO Name values for analysis and reporting, then converted to ISO 3-byte country codes for cross-walking into the map's visual abstraction. All runners with specified nations (whether by ISO code or name) were listed with their nation while all others were listed as 'United States.' For cross-walking locations to states, we identified 7,398 records where a state was given, ending in value such as ' AK',',AK', where 'AK' is 'Alaska'. This information was processed in creating Fig. 1 and Fig. 2.

Our dataset is complete, clean, and ready for further analysis.

```{r}
cherryblossom = na.omit(cherryblossom)
sum(is.na(cherryblossom))
```
```{r, include= TRUE}
head(cherryblossom)
```

# Results

## Race Results by Geographic Location
As visualized in Fig. 1, Results from analyzing geographic runner information identified that the nations with the fastest run times, represented by average, were from Tanzania, Morocco, Ethiopia, Kenya and Mexico. The nations with the slowest average run times were Ireland, Slovenia, Lebanon, Republic of Korea, and Jordan. 

| Top 5 Fastest Nations | Overall Race Completion Time |
|------|----------------------------------
| Tanzania | 46.067 |
| Morocco | 46.817 |
| Ethiopia | 46.983 |
| Kenya | 47.017 |
| Mexico | 47.317 |


| Top 5 Slowest Nations | Overall Race Completion Time |
|------|----------------------------------|
| Ireland | 121.42 |
| Slovenia | 108.25 |
| Lebanon | 103.95 |
| Republic of Korea | 100.733 |
| Jordan | 100.067 |

```{r International Results, include= TRUE, warning=F, fig.height=3.5}
library(pacman)
# loading all packages. The first 3 are for ISO codes and the usmap and map are for the literal maps
# also loading sqldf. Because of the volume of the data, sqldf was extremely faster than updating with conditionals in a for-loop;
# the time difference was roughly 5 seconds (sqldf) vs. 5 minutes (conditional for-loop)
p_load(rnaturalearth,rnaturalearthdata,countrycode,tidyverse,ggplot2,ggthemes,viridis,rgeos,usmap, sqldf, maps)

# These runners were all missing locations or had bad locations. We replaced them with what they had listed on file
# from previous races:
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Marla Hallacy",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Matthew Rogers",]$home = "Arlington VA"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Adri Jayaratne",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Marla Hallacy",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Eleanor Rathbone",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Emily Naden",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Grace L Chen",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Stephanie Rogers",]$home = "Aldie VA"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Sarah Garner",]$home = "	Arlington VA"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Maria Solomon",]$home = "Arlington VA"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Amy Iacopi",]$home = "San Francisco CA"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Sarah Watkins",]$home = "Washington DC"
cherryblossom[trimws(cherryblossom$name, "both", "[ \t\r\n]")=="Alissa Havens",]$home = "Washington DC"

cherryblossomX = cherryblossom
# the home column needed leading and trailing trimming
cherryblossomX$home = trimws(cherryblossomX$home, "both", "[ \t\r\n]")

# Standardizing countries to names where only codes were present. Although we revert them to codes for mapping,
# it is easier to work with the countries by name when coding
cherryblossomX[cherryblossomX$home =="AUS",]$home = "Australia"
cherryblossomX[cherryblossomX$home =="CAN",]$home = "Canada"
cherryblossomX[cherryblossomX$home =="COL",]$home = "Colombia"
cherryblossomX[cherryblossomX$home =="CZE",]$home = "Czech Republic"
cherryblossomX[cherryblossomX$home =="Fra",]$home = "France"
cherryblossomX[cherryblossomX$home =="GER",]$home = "Germany"
cherryblossomX[cherryblossomX$home =="IRE",]$home = "Ireland"
cherryblossomX[cherryblossomX$home =="JAP",]$home = "Japan"
cherryblossomX[cherryblossomX$home =="KOR",]$home = "Republic of Korea"
cherryblossomX[cherryblossomX$home =="LIT",]$home = "Lithuania"
cherryblossomX[cherryblossomX$home =="PER",]$home = "Peru"
cherryblossomX[cherryblossomX$home =="POL",]$home = "Poland"
cherryblossomX[cherryblossomX$home =="RSA",]$home = "South Africa"

cherryblossomX[cherryblossomX$home %in% c("ETH","Eth"),]$home = "Ethiopia"
cherryblossomX[cherryblossomX$home %in% c("KEN","Ken"),]$home = "Kenya"
cherryblossomX[cherryblossomX$home %in% c("MEX","Mex"),]$home = "Mexico"
cherryblossomX[cherryblossomX$home %in% c("RUS","Rus"),]$home = "Russian Federation"
cherryblossomX[cherryblossomX$home %in% c("ROM","Rom"),]$home = "Romania"
cherryblossomX[cherryblossomX$home %in% c("Us","Usa","Ny","MD"),]$home = "United States"
cherryblossomX[cherryblossomX$home %in% c("Bri","SCO"),]$home = "United Kingdom of Great Britain and Northern Ireland"

# dropping non-locational values
dropVars <- c("", "Lee","M")
cherryblossomX <- cherryblossomX[!(cherryblossomX$home %in% dropVars ),]

# Under the assumption US states have either leading spaces or commas (such as in Phoenix AZ/Phoenix,AZ/Phoenix, AZ for example),
# We recode states to 'United States' as follows in the SQL below. Also, Canadian states are recoded to 'Canada' in addition to
# the remaining nations being recoded to ISO standards. These were identified through manual, direct analysis. In the UK, one location
# with a code of LE was given for a city in Leicestershire. Finally, all remaining countries in the ISO list are provided and those not
# included are recorded to United States as it was specified in the documentation that American cities were not included with a national
# location (they, at best, included only state codes)
summary_table = sqldf("
      select 
      case when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' AL',' AK',' AZ',' AR',' CA',' CO',' CT',' DE',' DC',' FL',' GA',' HI',' ID',' IL',' IN',' IA',' KS',' KY',' LA',' ME',' MD',' MA',' MI',' MN',' MS',' MO',' MT',' NE',' NV',' NH',' NJ',' NM',' NY',' NC',' ND',' OH',' OK',' OR',' PA',' RI',' SC',' SD',' TN',' TX',' UT',' VT',' VA',' WA',' WV',' WI',' WY',' VI',',AL',',AK',',AZ',',AR',',CA',',CO',',CT',',DE',',DC',',FL',',GA',',HI',',ID',',IL',',IN',',IA',',KS',',KY',',LA',',ME',',MD',',MA',',MI',',MN',',MS',',MO',',MT',',NE',',NV',',NH',',NJ',',NM',',NY',',NC',',ND',',OH',',OK',',OR',',PA',',RI',',SC',',SD',',TN',',TX',',UT',',VT',',VA',',WA',',WV',',WI',',WY',',VI','afb','AFB') then 'United States'
           when trim(home) in ('California') then 'United States'
           when trim(home) in ('Indiana') then 'United States'
           when trim(home) in ('Wyoming') then 'United States'
           when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' CN',' ON',' NO',' SA',' BC',',CN',',ON',',NO','SA',',BC') then 'Canada'
           when substr( trim(home), length(trim(home))-3, length(trim(home)) ) in (' ONT',',ONT') then 'Canada'
           when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' UK',',UK',' LE',',LE') then 'United Kingdom of Great Britain and Northern Ireland'
           when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' PR',',PR') then 'Puerto Rico'
           when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' GE',',GE') then 'Germany'
           when trim(home) in ('Rep Of S.africa') then 'South Africa'
           when trim(home) in ('Scotland') then 'United Kingdom of Great Britain and Northern Ireland'
           when trim(home) in ('NewZealand') then 'New Zealand'
           when trim(home) in ('Russia') then 'Dominican Republic'
           when trim(home) in ('Dominican Republi') then 'Russian Federation'
           when trim(home) in ('Trinidad') then 'Trinidad and Tobago'
           when trim(home) in ('Tanzania') then 'Tanzania, United Republic of'
           when trim(home) in ('Bolivia') then 'Bolivia, Plurinational State of'
           when trim(home) not in ('Australia','Bolivia, Plurinational State of','Brazil','United Kingdom of Great Britain and Northern Ireland','Canada','Colombia','CzechRepublic','DominicanRepubli','DominicanRepublic','Dominican Republic','Czech Republic','South Africa','Puerto Rico','Ethiopia','France','Germany','Guatemala','Ireland','Japan','Kenya','Republic of Korea','Lithuania','Malta','Mexico','Morocco','NewZealand','New Zealand','Peru','Poland','PuertoRico','RepOfS.africa','Romania','Russian Federation','SouthAfrica','Trinidad and Tobago','Ukraine','UnitedKingdom','Uruguay','Tanzania, United Republic of','Slovenia','Jordan','Aruba', 'Angola', 'Aland Islands', 'Andorra', 'Argentina', 'American Samoa', 'French Southern and Antarctic Lands', 'Antigua and Barbuda', 'Austria', 'Burundi', 'Benin', 'Bonaire, Sint Eustatius and Saba', 'Burkina Faso', 'Bulgaria', 'Bahamas', 'Saint Barthelemy', 'Belize', 'Bolivia, Plurinational State of', 'Barbados', 'Bhutan', 'Botswana', 'Canada', 'Switzerland', 'China', 'Ivory Coast', 'Democratic Republic of the Congo', 'Cook Islands', 'Comoros', 'Costa Rica', 'Curacao', 'Cayman Islands', 'Czech Republic', 'Djibouti', 'Denmark', 'Algeria', 'Egypt', 'Western Sahara', 'Spain', 'Ethiopia', 'Fiji', 'France', 'Faroe Islands', 'Gabon', 'Georgia', 'Ghana', 'Guinea', 'Gambia', 'Equatorial Guinea', 'Grenada', 'Guatemala', 'Guam', 'Hong Kong', 'Heard Island and McDonald Islands', 'Croatia', 'Hungary', 'Isle of Man', 'British Indian Ocean Territory', 'Ireland', 'Iraq', 'Israel', 'Jamaica', 'Jordan', 'Kazakhstan', 'Kyrgyzstan', 'Kiribati', 'Saint Kitts and Nevis', 'Kuwait', 'Lebanon', 'Libya', 'Liechtenstein', 'Lesotho', 'Luxembourg', 'Macao', 'Morocco', 'Moldova, Republic of', 'Maldives', 'Marshall Islands', 'Mali', 'Myanmar', 'Mongolia', 'Mozambique', 'Montserrat', 'Mauritius', 'Malaysia', 'Namibia', 'Niger', 'Nigeria', 'Niue', 'Norway', 'Nauru', 'Oman', 'Panama', 'Peru', 'Palau', 'Poland', 'Portugal', 'Paraguay', 'French Polynesia', 'Reunion', 'Russian Federation', 'Saudi Arabia', 'Senegal', 'South Georgia and the South Sandwich Islands', 'Saint Helena, Ascension and Tristan da Cunha', 'Saint Helena, Ascension and Tristan da Cunha', 'Svalbard and Jan Mayen', 'Sierra Leone', 'San Marino', 'Saint Pierre and Miquelon', 'South Sudan', 'Suriname', 'Slovenia', 'Swaziland', 'Seychelles', 'Turks and Caicos Islands', 'Togo', 'Tajikistan', 'Turkmenistan', 'Tonga', 'Trinidad and Tobago', 'Turkey', 'Taiwan', 'Uganda', 'United States Minor Outlying Islands', 'United States', 'Holy See (Vatican City State)', 'Saint Vincent and the Grenadines', 'Virgin Islands, British', 'Vietnam', 'Wallis and Futuna', 'Kosovo', 'South Africa', 'Zimbabwe') then 'United States'
          else trim(home) end as home,
          runTime as runTime
      from cherryblossomX
      group by home
      ")

# getting the ISO codes from package 'maps'
dat <- iso3166
cherryblossom_country_runtimes <- sqldf("select d.a3, d.mapname, avg(x.runTime) as runTime from summary_table x join dat d on x.home = d.ISOname group by d.a3, d.mapname")

map <- ne_countries(scale = "medium", returnclass = "sf")
# Joining cross-walked ISO names to the map through the code names
runners_map <- merge(map, cherryblossom_country_runtimes, by.x = "adm0_a3", by.y = "a3", all = TRUE)

# global map
assoc_graph <- ggplot(data = runners_map) +
  geom_sf(aes(fill = runTime), 
          position = "identity") + 
  labs(fill='Average\nCompletion Time', caption = paste(
      'Figure 1:\n',
      'Global Race Completion Times', sep=''))  +
  scale_fill_viridis_c(option = "viridis") +
   ggtitle('Average Race Completion Time, by Country')

assoc_graph + theme_map()
```

As visualized in Fig. 2, the states represented by the fastest overall runners - based on average race completion time - are Nevada, Colorado, Delaware, Idaho, and New Mexico. The states with the slowest average race completion times are North Dakota, Kentucky, Hawaii, Oklahoma, and Oregon.

| Top 5 Fastest States | Overall Race Completion Time |
|------|----------------------------------|
| Nevada | 76.269 |
| Colorado | 82.400 |
| Delaware | 83.019 |
| Idaho | 83.157 |
| New Mexico | 83.208 |

| Top 5 Slowest States | Overall Race Completion Time |
|------|----------------------------------|
| North Dakota | 109.517 |
| Kentucky | 98.602 |
| Hawaii | 98.467 |
| Oklahoma | 97.837 |
| Oregon | 97.760 |

```{r National Results, include= TRUE, warning=F, fig.height=3.5}
cherryblossomStates = cherryblossom

# Recoding the states to a unified location.
states_summary_table = sqldf("
      select 
      case when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' AL',',AL') then 'AL'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' AK',',AK') then 'AK'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' AZ',',AZ') then 'AZ'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' AR',',AR') then 'AR'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' CA',',CA') then 'CA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' CO',',CO') then 'CO'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' CT',',CT') then 'CT'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' DE',',DE') then 'DE'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' FL',',FL') then 'FL'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' GA',',GA') then 'GA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' HI',',HI') then 'HI'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' ID',',ID') then 'ID'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' IL',',IL') then 'IL'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' IN',',IN') then 'IN'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' IA',',IA') then 'IA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' KS',',KS') then 'KS'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' KY',',KY') then 'KY'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' LA',',LA') then 'LA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' ME',',ME') then 'ME'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MD',',MD') then 'MD'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MA',',MA') then 'MA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MI',',MI') then 'MI'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MN',',MN') then 'MN'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MS',',MS') then 'MS'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MO',',MO') then 'MO'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' MT',',MT') then 'MT'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NE',',NE') then 'NE'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NV',',NV') then 'NV'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NH',',NH') then 'NH'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NJ',',NJ') then 'NJ'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NM',',NM') then 'NM'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NY',',NY') then 'NY'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' NC',',NC') then 'NC'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' ND',',ND') then 'ND'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' OH',',OH') then 'OH'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' OK',',OK') then 'OK'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' OR',',OR') then 'OR'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' PA',',PA') then 'PA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' RI',',RI') then 'RI'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' SC',',SC') then 'SC'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' SD',',SD') then 'SD'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' TN',',TN') then 'TN'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' TX',',TX') then 'TX'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' UT',',UT') then 'UT'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' VT',',VT') then 'VT'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' VA',',VA',' VI',',VI') then 'VA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' WA',',WA') then 'WA'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' WV',',WV') then 'WV'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' WI',',WI') then 'WI'
when substr( trim(home), length(trim(home))-2, length(trim(home)) ) in (' WY',',WY') then 'WY'
when trim(home) = 'Wyoming' then 'WY'
when trim(home) = 'California' then 'CA'
when trim(home) = 'Indiana' then 'IN'
          else NULL end as home,
          runTime as runTime
      from cherryblossomStates
      group by home
      ")

states_summary_table1 = na.omit(states_summary_table)

cherryblossom_sates_runtimes <- sqldf("select home as state, avg(runTime) as runTime from states_summary_table1 group by state")

# plotting the state-wide data. The only requirement is that the states have a column name 'state' filled with only the abbreviaitons
plot_usmap(data = cherryblossom_sates_runtimes, values = "runTime",  labels=FALSE)   +
  scale_fill_viridis_c(option = "viridis") + 
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) + 
  labs(fill='Average Race Completion Time', title = "Average Race Completion Time, by US State", caption = paste(
      'Figure 2:\n',
      'Statewide Race Completion Times', sep=''))
```

## Trends over time

The Cherry Blossom race has experienced growth since the year 2000 as indicated by the linear trend line in Fig. 3, below.

```{r, include = T, message = F, fig.height=3.5}
just_runTime = cherryblossom[c(1,6)]
runTime_by_year = aggregate(just_runTime, by = list(just_runTime$year), FUN = mean)
participation = aggregate(just_runTime, by = list(just_runTime$year), FUN = length)

# linear trend lines
participation_lm = lm(year ~ Group.1, data = participation)
runTime_lm = lm(runTime ~ year, data = runTime_by_year)
library(ggplot2)
p = ggplot(participation, aes(Group.1, runTime)) + 
  labs(title = 'Cherry Blossom Race Participation', caption = 'Figure 3: Cherry Blossom race participation with linear trend line') +
  xlab('Year') + ylab('Number of Participants') +
  theme(plot.title = element_text(hjust = .5), plot.caption = element_text(hjust = 0)) +
  scale_x_discrete(name = 'Year', limits = c(1999:2012))
p + geom_line() + geom_point() +
  geom_line(aes(y= participation_lm$fitted.values), color = 'blue', linetype = 'dotted') 
```

\pagebreak

The average race completion time for all ages and genders has also experienced a positive trend. The correlation between race participation and the average race completion time is 92.78%. This is indicated in Fig. 4, below.

```{r, include = T, message = F, fig.height=3.5}
p = ggplot(runTime_by_year, aes(year, runTime)) + 
  labs(title = 'Average Run Time 1999 - 2012', caption = 'Figure 4: Average race finish time with linear trend line') +
  xlab('Year') + ylab('Average Finish Time (minutes)') +
  theme(plot.title = element_text(hjust = .5), plot.caption = element_text(hjust = 0)) +
  scale_x_discrete(name = 'Year', limits = c(1999:2012))
p + geom_line() + geom_point() +
  geom_line(aes(y = runTime_lm$fitted.values), color = 'blue', linetype = 'dotted')
```
```{r}
cor(runTime_by_year$runTime, participation$runTime)
```

# Conclusion
These web-scraping methods proved useful for identifying the times of race completion per runner from year 1999 through 2012. From this information we were able to identify average race completion times, by nation and by state. However, documentation methodology and format varied over the years, which presented issues where data was required to be dropped - often as a result of missing information or incorrect information, such as children as young as 1 competing in the race - which impacted the results we were able to identify. National information was relatively reliable, but the state-level information was heavily under-represented. 

Because of the level of missing, unrecoverable information, we did not feel inferential analysis based on geographic location would have been useful. However, linear trajectory analysis proved useful when assessing racer count data over time; there is a clear indication race popularity - at least from the measure of race participant count - increased over the time spanning 1999 to 2012. Although with larger variance along the mean over time, there still appears to be a strong indication race completion time - measured in minutes - has also been increasing over time. That is, to say, during the same time period racer participation increased, race completion time increased as well.

\appendix


# Code

```
The code is cooool
```

